---
title: "HW5"
author: "Madelyn Maclaughlin"
date: "2024-11-04"
output: pdf_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

Load in the data:
```{r, message = FALSE}
library(tidyverse)
homicides <- read.csv('/Users/madelynmaclaughlin/Downloads/R_class/Ch_9/data/homicides.csv')
head(homicides)
```

I am creating the graph from choice 2. Let's try and reduce the data to what we need and organize it.
```{r}
homicides_dates <- homicides %>% 
  select(reported_date, city) %>% 
  filter(city == "Baltimore") %>% 
  mutate(reported_date = ymd(reported_date)) %>% 
  arrange(reported_date)
head(homicides_dates)
```

Sorting data into different groups based on season to add season labels to them, to separate them by color in the graph later. Then, I'm merging the two datasets into one so that I can graph everything all at once.
```{r}
summer <- homicides_dates %>% 
  mutate(months = month(reported_date, label = FALSE),
         years = year(reported_date)) %>% 
  filter(months %in% c(5, 6, 7, 8, 9, 10)) %>% 
  group_by(months, years) %>% 
  count() %>% 
  mutate(total_date = paste(years, months, sep = "-")) %>% 
  mutate(total_date = ym(total_date),
         season = "Summer")
  
winter <- homicides_dates %>% 
  mutate(months = month(reported_date, label = FALSE),
         years = year(reported_date)) %>% 
  filter(months %in% c(11, 12, 1, 2, 3, 4)) %>% 
  group_by(months, years) %>% 
  count() %>% 
  mutate(total_date = paste(years, months, sep = "-")) %>% 
  mutate(total_date = ym(total_date),
         season = "Winter")

w_and_s <- full_join(winter, summer, by = c("season", "total_date", "n", "years", "months")) %>% 
  mutate(season = as_factor(season))
```

Create a dataset of frequency for the geom_smooth function.
```{r}
h_freq <- homicides_dates %>% 
  mutate(month = round_date(reported_date, unit = "month")) %>% 
  group_by(month) %>% 
  count()
```


Begin graphing:

- span = 0.1 to make the geom_smooth line wiggly
- se = FALSE to remove error shadow
- used the seasons which had been converted to factors in scale_fill_manual to apply fill colors based on factor levels
- other code is mostly based on formatting axes, adding useful references, and working with the theme and legend.

```{r, fig.width = 7, fig.height = 2.5, message = FALSE}
ggplot() +
  geom_col(w_and_s, mapping = aes(x = total_date, y = n, fill = season), color = "gray50") +
scale_fill_manual(values = c("Winter" = "lightblue", "Summer" = "lightgray")) +
  geom_smooth(h_freq, mapping =  aes(x = month, y = n), se = FALSE, span = 0.1) +
  theme_dark() +
  labs(x = "Date", y = "Monthly Homicides", title = "Homicides in Baltimore, MD") +
  scale_y_continuous(breaks = c(0, 10, 20, 30, 40)) +
  geom_vline(xintercept = ymd("2015-04-12"), linewidth = 1.5, linetype = 8, color = "red") +
  geom_text(aes(x = ymd("2015-01-01"), y = 37, label = "Arrest of\nFreddie Gray"), 
            color = "white", hjust = 1) +
  theme(legend.position = "bottom", legend.title = element_blank())
```












